Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-05-22 manuscript/311 manuscript.tex
Nightly update
Created a new 5-year trend analysis and console file

\usepackage[margin = 1in]{geometry}
\usepackage{amsmath}
\allowdisplaybreaks
\usepackage{authblk}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{hyperref}\hypersetup{colorlinks=true, citecolor=blue}
\usepackage{parskip}
\usepackage{textgreek}
\usepackage{graphicx}
\usepackage{booktabs} % For better looking tables
\usepackage{siunitx} % For alignment of numbers
\usepackage{placeins}
\usepackage{caption}  % For custom captions
\usepackage{float}   % Required for the H specifier

@@

These (and other) challenges are constantly being addressed by the various agency open data managers
as well as the \href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology and Innovation} which
provides the technology support for the open data system.  
Many of the underlying datasets are updated daily, usually running 1-2 days behind from creation.
However, a number of datasets are updates monthly (such as restaurant inspections), and some are
not updated for one or more years (such as community board boundaries). This poses problems when 
merging these dataset with one another, as would be expected. Typically, the data must be
exported and manipulated in another program

@@

One result of these integrations, and other Agencies that use the 311 APIs, is that some of the data accuracies
and data consistency checks inherent in the 311 system are not applied uniformly, and in some cases not at all.
Errors often creep in from other Agency systems and are not easily corrected in the 311 application.

@@

Additionally, we extracted a 10-year dataset consisting of the years 2014-2023; some 27 million rows. Albeit with just a few columns, 
this very large dataset allows for long-term trend observations and confirms that the analysis using the 
2-year dataset used for the majority of the analysis.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{Yearly.png}
  \caption{SR counts by year for a 10-year period}
  \label{fig:yearly-counts}
\end{figure}

@@

This next chart shows the 10-year period by calendar month. Note that almost all the months before the COVID pandemic outbreak in Feb/Mar
of 2020 are ``below average'', while almost all the months after that period are ``above average'' months. And that trend of much higher usage
of the 311 system has continued into the early months of 2024.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{Monthly.png}
  \caption{SR counts by month for a 10-year period}
  \label{fig:monthly-counts}
\end{figure}

The daily tend over that same 10-year period is also interesting, as it shows both significant spikes
and significant dips; a noisy graph. While the spikes on 2020-08-04 (24,415 SRs) and 2020-08-05 (19,560 SRs) are COVID related, we are unable to 
ascertain any reason for the significant spike on Saturday, 2023-09-29 where 17,962 SRs were created.

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.35]{Daily.png}
  \caption{SR counts by day for a 10-year period}
  \label{fig:daily-counts}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{Calendar-Month.png}
  \caption{SR counts by calendar month for a 10-year period}
  \label{fig:calendar-months-counts}
\end{figure}

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{lS[table-format=7.0]S[table-format=6.0]} % 'l' for left-aligned, 'S' for siunitx number alignment
        \toprule
        \textbf{Month}     & \textbf{Count}   & \textbf{Count per Day} \\ 
        \midrule
        June  & 2454156 & 81805 \\ 
        July  & 2461525 & 79404 \\ 
        September & 2352822 & 78427 \\ 
        October & 2407238 & 77653 \\ 
        August & 2401731 & 77475 \\ 
        May  & 2334466 & 75305 \\ 
        November  & 2236037 & 74535 \\ 
        January   & 2271524 & 73275 \\ 
        December  & 2199765 & 70960 \\ 
        February  & 1996408 & 70795 \\ 
        March & 2174349 & 70140 \\ 
        April  & 2058169 & 68606 \\ 
       \bottomrule
    \end{tabular}
    \caption{Monthly counts sorted by counts-per-day}
    \label{tab:monthly_counts}
\end{table}

@@

\begin{figure}[htbp]
  \centering
	  \includegraphics[scale=0.65]{SRs_by_Agency.png}
	  \caption{SR counts by Agency and Cumulative Percentage}
	  \label{fig:SR_counts_by_Agency}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
	  \includegraphics[scale=0.65]{SRs_by_Complaint_Type.png}
	  \caption{Top 20 SR Complaints and Cumulative Percentage}
	  \label{fig:SR_complaints}
\end{figure}

@@

@@

\begin{itemize}
	\item Structural issues with the data. Is it in the expected format?
	\item Missing/blank data
	\item Data type correctness
	\item Invalid values
	\item Logical inconsistencies and inconsistent patterns in the data
	\item Accuracy and precision
	\item Redundant data  
\end{itemize}

\subsection{Structural Issues}

@@

In this dataset the writers discovered several structural issues.

	\subsubsection{Fields not in the Data Dictionary}
	The 311 SR \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
	 identifies 41 data columns (fields) along with some
	related information for each column. However, when you either download the data or explore it via the online portal,
	you will immediately note that there are 47 fields. The additional six fields are not addressed in the Data Dictionary, and show up on the Column Manager
	page as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer what the these @computed\_fields are, but it is not possible to know
	for certain. These fields are:
	\begin{itemize}
		\item zip\_codes
		\item community\_districts
		\item borough\_boundaries
		\item city\_council\_districts
		\item police\_precincts
		\item police\_precinct 
	\end{itemize}

@@

	\begin{figure}[htbp]
	  \centering
		  \includegraphics[scale=0.65]{computed_columns_screenshot.png}
		  \caption{Screenshot of data portal showing the ``@computed'' columns}
		  \label{fig:computed-columns}
	\end{figure}

@@

	\subsubsection{Verifying data types}
	Fortunately, both the Data Dictionary and the portal column manager do a good job of indicating the data type.
	As indicated by the red underline in Figure \ref{fig:computed-columns}, there is a small icon next to the column name 
	which indicates the data type of the column. The stylized capital ``T'' indicates text, the ``\#'' sign indicates 
	a numeric, a calendar icon indicates a date field, and the ``pin on a map'' symbol indicates a geospatial field.

@@

\subsection{Missing-Blank data}

@@

\begin{table}[ht]
\centering
\caption{Blank and N/A entries by field}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lrrrl}
	\toprule
	\textbf{Field} & \textbf{Total Empty} & \textbf{Percent Empty} & \textbf{NA Count} & \textbf{Blanks Only} \\
	\midrule
		taxi\_company\_borough & 6391341 & 99.94 & 0 & 6391341 \\
		road\_ramp & 6379080 & 99.75 & 0 & 6379080 \\
		vehicle\_type & 6376266 & 99.71 & 0 & 6376266 \\
		due\_date & 6370497 & 99.62 & 6370497 & 0 \\
		bridge\_highway\_direction & 6367626 & 99.57 & 0 & 6367626 \\
		bridge\_highway\_name & 6340197 & 99.14 & 0 & 6340197 \\
		bridge\_highway\_segment & 6340190 & 99.14 & 0 & 6340190 \\
		taxi\_pick\_up\_location & 6329954 & 98.98 & 0 & 6329954 \\
		facility\_type & 5966588 & 93.30 & 0 & 5966588 \\
		landmark & 2635736 & 41.22 & 0 & 2635736 \\
		intersection\_street\_1 & 2143847 & 33.52 & 0 & 2143847 \\
		intersection\_street\_2 & 2140707 & 33.48 & 0 & 2140707 \\
		cross\_street\_1 & 1846594 & 28.88 & 0 & 1846594 \\
		cross\_street\_2 & 1846008 & 28.87 & 0 & 1846008 \\
		location\_type & 798775 & 12.49 & 0 & 798775 \\
		bbl & 768109 & 12.01 & 0 & 768109 \\
		city & 332086 & 5.19 & 0 & 332086 \\
		street\_name & 280463 & 4.39 & 0 & 280463 \\
		incident\_address & 280268 & 4.38 & 0 & 280268 \\
		closed\_date & 248839 & 3.89 & 248839 & 0 \\
		resolution\_description & 132333 & 2.07 & 0 & 132333 \\
		zip\_codes & 123167 & 1.93 & 0 & 123167 \\
		borough\_boundaries & 101453 & 1.59 & 0 & 101453 \\
		community\_districts & 101445 & 1.59 & 0 & 101445 \\
		city\_council\_districts & 101445 & 1.59 & 0 & 101445 \\
		police\_precincts & 101445 & 1.59 & 0 & 101445 \\
		police\_precinct & 101416 & 1.59 & 0 & 101416 \\
		latitude & 99778 & 1.56 & 0 & 99778 \\
		longitude & 99778 & 1.56 & 0 & 99778 \\
		location & 99778 & 1.56 & 0 & 99778 \\
		x\_coordinate\_state\_plane & 99668 & 1.56 & 0 & 99668 \\
		y\_coordinate\_state\_plane & 98822 & 1.55 & 0 & 98822 \\
		resolution\_action\_updated\_date & 91399 & 1.43 & 91399 & 0 \\
		incident\_zip & 83220 & 1.30 & 2 & 83218 \\
		descriptor & 74645 & 1.17 & 0 & 74645 \\
		address\_type & 38655 & 0.60 & 0 & 38655 \\
		unique\_key & 0 & 0.00 & 0 & 0 \\
		agency & 0 & 0.00 & 0 & 0 \\
		agency\_name & 0 & 0.00 & 0 & 0 \\
		complaint\_type & 0 & 0.00 & 0 & 0 \\
		status & 0 & 0.00 & 0 & 0 \\
		community\_board & 0 & 0.00 & 0 & 0 \\
		borough & 0 & 0.00 & 0 & 0 \\
		open\_data\_channel\_type & 0 & 0.00 & 0 & 0 \\
		park\_facility\_name & 0 & 0.00 & 0 & 0 \\
		park\_borough & 0 & 0.00 & 0 & 0 \\
		created\_date & 0 & 0.00 & 0 & 0 \\
	\bottomrule
\end{tabular}
}
\end{table}

@@

Here is a graphic depiction of total empty (blank + N/As) for each fields. You can see the natural grouping into the Most, Some, and Few categories.

\begin{figure}[htbp]
  \centering
	  \includegraphics[scale=0.7]{BlankFields.png}
	  \caption{Number and Percentage of Empty Entries}
	  \label{fig:blank_fields}
\end{figure}

\subsection{Data type correctness}
All data fields were subjected to a test of ``correctness'' according to their ``type'' as specified
in the Data Dictionary.  Of the nearly 300 million data fields to check (6.5 million rows, 47 fields per row), there were only two  values that failed to test for the
proper data type; those being two text entries in the incident\_zip filed (``na'', and N/A). All date fields were dates. All number fields were numeric. All text
fields were character.

\subsection{Checking for Valid Values}
Testing for valid and invalid values revealed several fairly serious issues associated with select fields. Accordingly, an analyst would need to be
very diligent if using these fields for analysis. In all likelihood, the rows with the invalid field values would likely need to be removed from the
analysis.

There were a variety of tests for validity that were applied individually to specific fields. Here are the ``good'' results:

\begin{itemize}
		\item Latitude and Longitude fields were tested to ensure all fell within the geographic boundaries of the City of New York. All were.
		\item The unique\_key field was in fact unique.
		\item Many fields have a domain of acceptable values. Often these values are determined from common usage or by examining
		larger datasets. Unfortunately, these domains of acceptable values are not specified in the Data Dictionary. These fields all tested as
		compliant with their domain of acceptable values:
			\begin{itemize}
				\item address\_type
				\item status
				\item borough, borough\_boundaries, \& park\_borough 
				\item data\_channel
				\item vehicle\_type
				\item city\_council\_district
			\end{itemize}
\end{itemize}

 Unfortunately some fields did prove to be quite problematic when comparing the values to the domain of legal values.
 For example, all zip codes (two fields: zip\_codes and incident\_zip) should be valid as defined by the USPS database
 which contains over 44,000 valid zip codes.

@@

\begin{figure}[htbp]
  \centering
	  \includegraphics[scale=0.7]{invalid_incident_zip.png}
	  \caption{Invalid incident-zip by Agency}
	  \label{fig:invalid_incident_zip}
\end{figure}

A curious case also exists when examining the two nearly identical fields - police\_precincts and police\_precinct.

\subsection{Logical inconsistencies }

\subsection{Accuracy and precision}

\subsection{Redundant data}

@@

\subsection{Case Study I - Studying Noise Complaints by NYC Zip Code}

Invalid values?
Logical inconsistencies? Concerning patterns?
Redundant columns?

Invalid data: closed data earlier than creation date;
latitude/longitude not in NYC; invalid zip code;

Inconsistent data: upper/lower cases; extra spaces; 
open text entries: provide enough major categories and sub categories;

>>>>

\usepackage[margin = 1in]{geometry}
\usepackage{amsmath}
\allowdisplaybreaks
\usepackage{authblk}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{hyperref}\hypersetup{colorlinks=true, citecolor=blue}
\usepackage{parskip}
\usepackage{textgreek}
\usepackage{graphicx}
\usepackage{booktabs} % For better looking tables
\usepackage{siunitx} % For alignment of numbers
\usepackage{placeins}
\usepackage{caption}  % For custom captions
\usepackage{float}   % Required for the H specifier
\usepackage{adjustbox}

@@

These (and other) challenges are constantly being addressed by the various agency open data managers
as well as the \href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology and Innovation (OTI)} which
provides the technology support for the open data system.  
Many of the underlying datasets are updated daily, usually running 1-2 days behind from creation.
However, a number of datasets are updates monthly (such as restaurant inspections), and some are
not updated for one or more years (such as community board boundaries). This poses problems when 
merging these dataset with one another, as would be expected. Typically, the data must be
exported and manipulated in another program

@@

One result of these integrations, and other Agencies that use the 311 APIs, is that some of the data accuracies
and data consistency checks inherent in the 311 system are not applied uniformly, and in some cases not at all.
Errors often creep in from other Agency systems and are not easily corrected in the 311 application itself,
but rather require correction at the source Agency.

@@

Additionally, we extracted two longer-term datasets, a 5-year look (2019-2023) and a 10-year look (2014-2023) ; some 14  and 27 million rows respectively, albeit with just a few columns. 
These larger datasets afford an opportunity for long-term trend observations, confirming the analysis using the 2-year dataset. The 2022-2023 dataset
was used exclusively for the detailed, field-level analysis that comprises the majority of this effort.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{Yearly.png}
  \caption{SR counts by year for a 10-year period}
  \label{fig:yearly-counts}
\end{figure}

@@

This below chart shows the 10-year period by calendar month. Note that almost all the months before the COVID pandemic outbreak in Feb/Mar
of 2020 are ``below average'', while most of the months after that period are ``above average'' months. And that trend of higher usage
of the 311 system has continued into the early months of 2024.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{Monthly.png}
  \caption{SR counts by month for a 10-year period}
  \label{fig:monthly-counts}
\end{figure}

A quick look at the corresponding  daily tend over that same 10-year period is also interesting, as it shows both significant spikes
and significant dips; a noisy graph. While the spikes on 2020-08-04 (24,415 SRs) and 2020-08-05 (19,560 SRs) are COVID related, we are unable to 
ascertain any reason for the significant spike on Saturday, 2023-09-29 where 17,962 SRs were created.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Daily.png}
  \caption{SR counts by day for a 10-year period}
  \label{fig:daily-counts}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Calendar-Month.png}
  \caption{SR counts by calendar month for a 10-year period}
  \label{fig:calendar-months-counts}
\end{figure}

@@

\begin{table}[ht]
    \centering
    \small
    \begin{tabular}{lS[table-format=7.0]S[table-format=6.0]} % 'l' for left-aligned, 'S' for siunitx number alignment
        \toprule
        \textbf{Month}     & \textbf{Count}   & \textbf{Count per Day} \\ 
        \midrule
        June  & 2454156 & 81805 \\ 
        July  & 2461525 & 79404 \\ 
        September & 2352822 & 78427 \\ 
        October & 2407238 & 77653 \\ 
        August & 2401731 & 77475 \\ 
        May  & 2334466 & 75305 \\ 
        November  & 2236037 & 74535 \\ 
        January   & 2271524 & 73275 \\ 
        December  & 2199765 & 70960 \\ 
        February  & 1996408 & 70795 \\ 
        March & 2174349 & 70140 \\ 
        April  & 2058169 & 68606 \\ 
       \bottomrule
    \end{tabular}
    \caption{Monthly counts sorted by counts-per-day}
    \label{tab:monthly_counts}
\end{table}

@@

\begin{figure}[htbp]
  \centering
	  \includegraphics[width=\textwidth]{SRs_by_Agency.png}
	  \caption{SR counts by Agency and Cumulative Percentage}
	  \label{fig:SR_counts_by_Agency}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
	  \includegraphics[width=\textwidth]{SRs_by_Complaint_Type.png}
	  \caption{Top 20 SR Complaints and Cumulative Percentage}
	  \label{fig:SR_complaints}
\end{figure}

@@

There are also some rather curious and unusual Service Requests in the 311 data: 
tanning, tattooing, trans fat, unsanitary pigeon condition, illegal animal kept as pet, harboring bees/wasps, and radioactive material. 
New Yorkers certainly live interesting lives.

@@

\begin{itemize}
	\item Structural issues with the data. Is it in the expected format?
	\item Data type correctness
	\item Missing/blank data
	\item Invalid values
	\item Logical inconsistencies and inconsistent patterns in the data
	\item Accuracy and precision
	\item Redundant data  
\end{itemize}

\section{Structural Issues}

@@

\textbf{Fields not in the Data Dictionary} The 311 SR \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
 identifies 41 data columns (fields) along with some
related information for each column. However, when you either download the data or explore it via the online portal,
you will immediately note that there are 47 fields. The additional six fields are not addressed in the Data Dictionary, and show up on the Column Manager
page as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer what the these @computed\_fields are, but it is not possible to know
for certain. These fields are:
\begin{itemize}
	\item zip\_codes
	\item community\_districts
	\item borough\_boundaries
	\item city\_council\_districts
	\item police\_precincts
	\item police\_precinct 
\end{itemize}

@@

\begin{figure}[htbp]
  \centering
	  \includegraphics[scale = 0.7]{computed_columns_screenshot.png}
	  \caption{Screenshot of data portal showing the ``@computed'' columns}
	  \label{fig:computed-columns}
\end{figure}

@@

\section{Verifying data types}
Fortunately, both the Data Dictionary and the portal column manager do a good job of indicating the data type.
As indicated by the red underline in Figure \ref{fig:computed-columns}, there is a small icon next to the column name 
which indicates the data type of the column. The stylized capital ``T'' indicates text, the ``\#'' sign indicates 
a numeric, a calendar icon indicates a date field, and the ``pin on a map'' symbol indicates a geospatial field.

@@

\section{Blank \& N/A data}

@@

\begin{table}[ht]
\centering
\caption{Blank and N/A entries by field}
\resizebox{0.8\linewidth}{!}{%
\tiny
\begin{tabular}{lrrrl}
	\toprule
	\textbf{Field} & \textbf{Total Empty} & \textbf{Percent Empty} & \textbf{N/A Count} \\
	\midrule
		taxi\_company\_borough & 6391341 & 99.94 & 0 \\
		road\_ramp & 6379080 & 99.75 & 0 \\
		vehicle\_type & 6376266 & 99.71 & 0 \\
		due\_date & 6370497 & 99.62 & 6370497 \\
		bridge\_highway\_direction & 6367626 & 99.57 & 0 \\
		bridge\_highway\_name & 6340197 & 99.14 & 0 \\
		bridge\_highway\_segment & 6340190 & 99.14 & 0 \\
		taxi\_pick\_up\_location & 6329954 & 98.98 & 0 \\
		facility\_type & 5966588 & 93.30 & 0 \\
		landmark & 2635736 & 41.22 & 0 \\
		intersection\_street\_1 & 2143847 & 33.52 & 0 \\
		intersection\_street\_2 & 2140707 & 33.48 & 0 \\
		cross\_street\_1 & 1846594 & 28.88 & 0 \\
		cross\_street\_2 & 1846008 & 28.87 & 0 \\
		location\_type & 798775 & 12.49 & 0 \\
		bbl & 768109 & 12.01 & 0 \\
		city & 332086 & 5.19 & 0 \\
		street\_name & 280463 & 4.39 & 0 \\
		incident\_address & 280268 & 4.38 & 0 \\
		closed\_date & 248839 & 3.89 & 248839 \\
		resolution\_description & 132333 & 2.07 & 0 \\
		zip\_codes & 123167 & 1.93 & 0 \\
		borough\_boundaries & 101453 & 1.59 & 0 \\
		community\_districts & 101445 & 1.59 & 0 \\
		city\_council\_districts & 101445 & 1.59 & 0 \\
		police\_precincts & 101445 & 1.59 & 0 \\
		police\_precinct & 101416 & 1.59 & 0 \\
		latitude & 99778 & 1.56 & 0 \\
		longitude & 99778 & 1.56 & 0 \\
		location & 99778 & 1.56 & 0 \\
		x\_coordinate\_state\_plane & 99668 & 1.56 & 0 \\
		y\_coordinate\_state\_plane & 98822 & 1.55 & 0 \\
		resolution\_action\_updated\_date & 91399 & 1.43 & 91399 \\
		incident\_zip & 83220 & 1.30 & 2 \\
		descriptor & 74645 & 1.17 & 0 \\
		address\_type & 38655 & 0.60 & 0 \\
		unique\_key & 0 & 0.00 & 0 \\
		agency & 0 & 0.00 & 0 \\
		agency\_name & 0 & 0.00 & 0 \\
		complaint\_type & 0 & 0.00 & 0 \\
		status & 0 & 0.00 & 0 \\
		community\_board & 0 & 0.00 & 0 \\
		borough & 0 & 0.00 & 0 \\
		open\_data\_channel\_type & 0 & 0.00 & 0 \\
		park\_facility\_name & 0 & 0.00 & 0 \\
		park\_borough & 0 & 0.00 & 0 \\
		created\_date & 0 & 0.00 & 0 \\
	\bottomrule
\end{tabular}
}
\end{table}

@@

Here is a graphic depiction of total empty (blank \& N/As) for each fields. You can see the natural grouping into the Most, Some, and Few categories.

\begin{figure}[H]
  \centering
	  \includegraphics[width=\textwidth]{BlankFields.png}
	  \caption{Number and Percentage of Empty Entries}
	  \label{fig:blank_fields}
\end{figure}

\section{Data type correctness}

All data fields were subjected to a test of ``correctness'' according to their ``type'' as specified
in the Data Dictionary.  Of the nearly 300 million data fields to check (6.5 million rows, 47 fields per row), there were only two  values that failed to test for the
proper data type; those being two text entries in the incident\_zip filed (``na'', and N/A). All date fields were dates. All number fields were numeric. All text
fields were character.

 \section{Validating data}
 While the dataset may have fields populated with data that appears correct, it is necessary to inspect those fields to determine if the data is indeed valid. For
 example a date field that contains the value ``February 30, 2024'' is clearly not valid. Nor is a latitude of +95 deg north.

 In this study, testing for invalid values revealed several serious issues associated with select fields. Accordingly, an analyst would need to be
extra diligent if using these fields for analysis. In all likelihood, rows with invalid field values would likely need to be removed from the
analysis dataset.

Some fields can be validated against either a acceptable values, often drawn from historical context. In this study we looked at a 10-year
dataset to ascertain a domain for values what do not have a specific reference source. We would recommend that these domains be published
in the Data Dictionary. Here are some results.

\begin{itemize}
	\item Latitude and Longitude fields were tested to ensure all fell within the geographic boundaries of the City of New York. All were.
	\item The unique\_key field was in fact unique.
	\item Many fields have a domain of acceptable values. Often these values are determined from common usage or by examining
	larger historical datasets. Unfortunately, these domains of acceptable values are not specified in the Data Dictionary. These fields all tested as
	compliant with their domain of acceptable values:
		\begin{itemize}
			\item address\_type
			\item status
			\item borough, borough\_boundaries, \& park\_borough 
			\item data\_channel
			\item vehicle\_type
			\item city\_council\_district
		\end{itemize}
\end{itemize}

	\subsection{Issues with Zip Codes}
	 Unfortunately some fields did prove to be quite problematic when comparing the values to the domain of legal values.
	 For example, all zip codes (two fields: zip\_codes and incident\_zip) should be valid as defined by the USPS database
	 which contains over 44,000 valid zip codes.

@@

	\begin{figure}[H]
	  \centering
		  \includegraphics[width=\textwidth]{invalid_incident_zip.png}
		  \caption{Invalid incident-zip by Agency}
		  \label{fig:invalid_incident_zip}
	\end{figure}

		\subsubsection{Case Study: Noise Complaints by Zip Code}
		\textbf{Scenario:} The NYC Office of Nightlife (ONL) wants to know ``What are the top 10 zip codes for Noise Complaints (all 8 types) over the last two years?''
		The goal is to assess the impact of the recent NYC effort looking to promote a safe and vibrant nightlife scene in NYC while seeking to ease
		strained relations between bar and club owners.

		It may come as a surprise to non-NYC residents that dancing at bars and restaurants has been illegal since 1926 (during prohibition), and was only just repealed in 2018.

		On the surface this would seem like a simple effort.  The NYC Open Data Portal allows for selection of a timeframe (2022-2023), complaint-type
		begins with ``Noise'', and group by Zip code, sorted descending by count.  You can do this analysis and grouping right inside the NYC Open Data Portal. \textit{Voila!}

	\begin{figure}[H]
	    		\centering
	    		\includegraphics[scale = 0.6]{zipcode_casestudy.png}
	    		\caption{Top 10 zip codes for Noise complaints (Zip Codes)}
	    	\label{fig:casestudy1-zipcodes}
		\end{figure}

		However, this analysis uses the zip\_codes field, one of the (six) computed fields that has shown to have validity problems. If we repeat the analysis with the 
		incident\_zip field instead of the zip\_codes field, we obtain a different set of zip code results:

		\begin{figure}[H]
	    		\centering
	    		\includegraphics[scale = 0.6]{incident_zip_casestudy.png}
	    		\caption{Top 10 zip codes for Noise complaints(Incident Zip)}
	    	\label{fig:casestudy1-incident-zip}
		\end{figure}

	Let's subject these two data sets to validation against the US Postal Service zip code database.

	\begin{table}[H]
	\centering
	\footnotesize
	\caption{Comparison of Top Ten Zip Codes Lists}
	\begin{tabular}{lrr@{\hskip 0.5cm}|@{\hskip 0.5cm}lrr}
	\toprule
	\multicolumn{3}{c}{\textbf{zip\_codes}} & \multicolumn{3}{c}{\textbf{incident\_zip}} \\
	\cmidrule(r){1-3} \cmidrule(l){4-6}
	\textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} & \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} \\
		\midrule
		11275 & 104,556 & FALSE & 10466 & 104,562 & TRUE \\
		12420 & 27,503 & TRUE & 10023 & 27,972 & TRUE \\
		12428 & 26,564 & TRUE & 10031 & 25,548 & TRUE \\
		10935 & 25,508 & FALSE & 10457 & 25,066 & TRUE \\
		10934 & 23,448 & FALSE & 10453 & 24,752 & TRUE \\
		10931 & 22,381 & TRUE & 10456 & 24,751 & TRUE \\
		10930 & 22,121 & TRUE & 10452 & 22,527 & TRUE \\
		17613 & 21,963 & FALSE & 10025 & 21,705 & TRUE \\
		10936 & 21,707 & FALSE & 10458 & 21,689 & TRUE \\
		11606 & 21,435 & FALSE & 10032 & 20,622 & TRUE \\
		\bottomrule
	\end{tabular}
	\end{table}

	As indicated, six out of ten zip\_codes are invalid, which corresponds closely with what is observed in the overall dataset (58\%). Whereas the incident\_zip
	field is completely valid, again in-line with the overall incident\_zip validation percentage (99.04\%)
	Perhaps more curious than the differing zip codes on the two lists, is the fact that the numerical counts are nearly identical;  the overall counts are quite close. This is curious since
	one field is computed while the other is via data entry; the computed protocol is again the suspect.

	\subsection{Issues with Police Precincts}
	A curious case also exists when examining the two nearly identical fields - police\_precincts and police\_precinct. Both of those fields are among
	the ``computed'' fields in the dataset. 
	Using  \href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}{NYPD Precinct listings} it's possible to determine the valid
	police precincts; they are generally numeric or have numeric representations in the dataset.

	However, what we find is that both fields police\_precinct and police\_precincts  have  35\% invalid entries.
	Unfortunately, they're not the same invalid entries.

	For the police\_precincts field, there are 2,171,864 invalid entries (35\%)., a consequential number of errors. Similarly for the police\_precinct field,
	 there are 2,171,778 invalid entries ( also 35\%), however they are not the same counts for the two similar fields.

	The top ten (by count) of invalid precincts are:

	\begin{table}[ht]
	\footnotesize
	\centering
		\begin{tabular}{ccccc}
		\toprule
		\textbf{Invalid Precinct} & \textbf{police\_precinct count} & \textbf{police\_precincts count} & \textbf{Delta}\\
		\midrule
			62 & 151808 & 151720 & 88 \\
			72 & 148881 & 148796 & 85 \\
			67 & 133171 & 133115 & 56 \\
			64 & 111236 & 111221 & 15 \\
			60 & 105252 & 105317 & -65 \\
			65 & 103334 & 103306 & 28 \\
			53 & 100097 & 100074 & 23 \\
			73 & 97746 & 97755 & -9 \\
			68 & 94204 & 94254 & -50 \\
			66 & 93654 & 93848 & -194 \\
		\bottomrule
		\end{tabular}
	\caption{Comparison of the fields police\_precinct and police\_precincts}
	\label{tab:comparison-precincts-diff}
	\end{table}

	A graphic representation of the invalid police\_precincts plotted by Agency show a near perfect match to the
	distribution of the entire 6.4 million SRs. Note that the ``big six'' Agencies account for 90\% of these invalid precinct errors, identical
	to the overall distribution of all SRs. This strongly suggests that this is a systemic problem, almost certainly
	caused by the computation method applied for these two fields.

	\begin{figure}[H]
	  \centering
		  \includegraphics[width=\textwidth]{invalid_police_precincts.png}
		  \caption{Invalid NYPD precincts by Agency}
		  \label{fig:invalid_police_precincts_zip}
	\end{figure}

	\subsection{Issues with Community Boards}
	Community Boards are an important aspect of NYC government. Community boards are the most local, grassroots form of City government, 
	and the connection between communities and elected officials and City agencies. They are organized by the five Boroughs of NYC and represented
	in this dataset as ``\#\#-Borough'', e.g. ``10 Bronx''. As such, they play an important role in the quality of life for all New Yorkers. There are 59 community boards 
	throughout the City. There are 12 Community Districts in the Bronx, 18 in Brooklyn, 12 in Manhattan, 14 in Queens and 3 in Staten Island. The Community Boards are
	used frequently as a way to measure City services throughout the five boroughs, so correctness is important. NYC Borough Presidents in particular are focused
	on the services offered at the Community Board level.

	In the 2022-2023 dataset there are 27,276 invalid community\_board entries which represents 0.43\% of non-blank data. There are a total of 12 different invalid community boards.
	These include:

		\begin{table}[H]
		\centering
		\small
		\caption{Top Ten Invalid 'community\_board' Values}
		\begin{tabular}{rlr}
		\toprule
		\textbf{Rank} & \textbf{Invalid CB} & \textbf{Count} \\
			\midrule
				1 & 64 MANHATTAN & 7144 \\
				2 & 83 QUEENS & 5133 \\
				3 & 55 BROOKLYN & 3327 \\
				4 & 81 QUEENS & 3180 \\
				5 & 80 QUEENS & 2407 \\
				6 & 26 BRONX & 2364 \\
				7 & 28 BRONX & 1128 \\
				8 & 82 QUEENS & 1054 \\
				9 & 95 STATEN ISLAND & 515 \\
				10 & 27 BRONX & 514 \\
			\bottomrule
		\end{tabular}
		\end{table}

		The distribution of invalid Community Boards by Agency is not consistent with the overall SR Agency distribution. This indicates that there are likely
		specific issues at some key Agencies, in this case Taxi \& Limo Commission (TLC), Parks \& Recreation, etc. It's not a large error, but the division of services
		(and complaints) by Community Board is a well-tracked measure.

		\begin{figure}[H]
	  \centering
		  \includegraphics[scale = 0.65]{invalid_community_boards.png}
		  \caption{Invalid Community Board by Agency}
		  \label{fig:invalid_community_boards}
	\end{figure}

	\subsection{Issues with Community Districts}
	Another of the computed fields is community\_districts. Community Districts are the boundaries for the Community Boards, but unlike how Community Boards are
	an arm of the government of New York City, the Community District is used by the Department of City Planning (DCP) for purposes of environmental, socio-economic, and
	demographic purposes. DCP is the NYC''s primary land use agency and is instrumental in the City's physical framework.  It is a geographical division rather
	than a local government division. And like Community Boards, the Community Districts are a frequently used means to breakdown zoning, housing, community facilities,
	and waterfront/open/public spaces. It is also a well-tracked measure.

	Due to how the community\_district data is formatted, it is not possible to establish validity. However, it is possible to determine that the dataset contains
	72 unique entries, while there are only 59 valid Community Districts.

\section{Logical inconsistencies }

closed before created (negative duration)
closed exactly as created (zero duration)
closed in the future

\subsubsection Case Study: Homeless Person Assistance

due\_date before created
long resolution\_action\_updated\_date

\section{Accuracy and precision}
There is one area in particular where the question of precision vs. accuracy immediately arises, and that is with the Latitude and Longitude fields.  Both the 
Longitude and Latitude fields are expresses as a 14-decimal number, e.g. 40.86769186022511 (also the Location field which is a straight concatenation of
latitude and longitude). Given that 1 degree of latitude at the equator is equal to 111.044736 kilometers, the ``1'' at the end of that number represents
approximate 1.1104 nanometer or 1/1,000,000,000 of a meter. A DNA molecule is approximately 2nm in width.

Clearly the representation of the Latitude and Longitude fields are a classic case of 14-digit precision, 
but limited accuracy. It is more likely that the Lat/Long values are accurate to the 5\textsuperscript{th} decimal place, about 3.64 feet; even that might be
overstating things.

@@

\section{Redundant \& Duplicate data}
Shrinking file size by deleting these 11 redundant fields:
     1 - agency\_name 
     2 - park\_borough 
     3 - borough\_boundaries 
     4 - location 
     5 - intersection\_street\_1 
     6 - intersection\_street\_2 
     7 - police\_precinct 
     8 - duration 
     9 - postClosedUpdateDuration 
     10 - translated\_borough\_boundaries 
     11 - zip\_codes 
Original size: 3.2 Gb 
Size after removing redundant columns: 2.8 Gb 
Potential size reduction: 395.3 Mb or 12.3
